---
title: Exploring complex networks with the `ICON` R package
authors:
  - name: Raoul R. Wadhwa
    department: Cleveland Clinic Lerner College of Medicine
    affiliation: Case Western Reserve University
    location: Cleveland, OH 44195
    email: raoul.wadhwa@case.edu
  - name: Jacob G. Scott
    department: Translational Hematology \& Oncology Research
    affiliation: Lerner Research Institute, Cleveland Clinic
    location: Cleveland, OH 44195
    email: ScottJ10@ccf.org
abstract: |
  We introduce `ICON`, an R package that contains `r nrow(ICON::ICON_data)` complex network datasets in a standard edgelist format.
  All provided datasets have associated citations and have been indexed by the Colorado Index of Complex Networks - also referred to as ICON.
  In addition to supplying a large and diverse corpus of useful real-world networks, `ICON` also implements an S3 generic to work with the `network` and `ggnetwork` R packages for network analysis and visualization, respectively.
  Sample code in this report also demonstrates how `ICON` can be used in conjunction with the `igraph` package.
  Currently on version $0.4.0$, `ICON` has been accepted by the Comprehensive R Archive Network.
  It is our hope that `ICON` will serve as a standard corpus for complex network research and prevent redundant work that would be otherwise necessary by individual research groups.
  The open-source code for `ICON` and for this reproducible report can be found at \url{https://github.com/rrrlw/ICON}.
keywords:
  - complex networks
  - R programming language
bibliography: ICONrefs.bib
csl: num_style.csl
output: rticles::arxiv_article
---

# Introduction

Empirical study of complex networks requires real-world data to validate theoretical results.
A large, diverse corpus of networks often proves useful given the many shapes and sizes that complex networks assume.[@corpus1;@corpus2]
To our knowledge, the [Colorado Index of Complex Networks (ICON)](https://icon.colorado.edu) hosts the largest curated index of real-world complex networks, with metadata and links to over 5,000 networks as of this writing.[@icon]
However, heterogeneity in data format, access, and availability limit how easily users can take advantage of this incredible resource.
A central repository containing a large corpus of ICON-indexed networks in standard format would thus provide a useful service for network science researchers, who would avoid the tedious task of data format conversion prior to analysis.
Here, we introduce the `ICON` R package as such a solution, providing a large and diverse corpus of real-world networks and tools to work with existing network analysis and visualization R packages.

# Implementation details

\texttt{ICON} v$0.4.0$ is a package for the R programming language hosted on the Comprehensive R Archive Network.[@rlang]
It strongly depends on the `network` ($\geq 1.16$) and `utils` ($\geq 3.6.1$) R packages and weakly depends on the `covr` ($\geq 3.5$), `ggnetwork` ($\geq 0.5$), `ggplot2` ($\geq 3.3$), `knitr` ($\geq 1.30$), `testthat` ($\geq 2.3$), and `rmarkdown` ($\geq 2.4$) R packages.[@rlang;@network-pkg;@covr-pkg;@ggnetwork-pkg;@ggplot2-pkg;@knitr-pkg;@testthat-pkg;@rmarkdown-pkg]
Throughout the development process, the `devtools` R package was heavily used.[@devtools-pkg]
This report is fully reproducible with code (GPL-3 license) available at \url{https://github.com/rrrlw/ICON}.
Full reproduction will require installation of the strong and weak dependencies listed above.

## Installation

The stable version of `ICON` (currently v$0.4.0$) can be downloaded from CRAN, while the development version can be downloaded from the package's [GitHub repository](https://github.com/rrrlw/ICON) using the `remotes` package.[@remotes-pkg]
Both options are demonstrated in the following code chunk.

```{r install, eval=FALSE}
# install stable version from CRAN
install.packages("ICON")

# install development version from GitHub
remotes::install_github("rrrlw/ICON", build_vignettes = TRUE)
```

## Complex network datasets (and hosting)

## The ICON class and network S3 generic

# Use cases

Before starting with the use cases, the following code chunk will load the appropriate libraries and download the sample dataset.

```{r use-setup,message=FALSE,results='hide'}
# load necessary libraries
library("ICON")
library("network")
library("ggnetwork")
library("ggplot2")
library("igraph")

# download sample dataset
get_data("seed_disperse_beehler")
```

## With the \texttt{network} R package

```{r network-case,message=FALSE,results='hide',fig.width=3,fig.height=3}
# setup using as.network generic
coerced <- as.network(seed_disperse_beehler)

# plot with ggnetwork
ggplot(coerced, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(alpha = 0.25) +
  geom_nodes() +
  theme_blank()
```

## With the \texttt{igraph} R package

```{r igraph-case,message=FALSE,results='hide',fig.width=4,fig.height=4}
# plot network using igraph
my_graph <- graph_from_edgelist(as.matrix(seed_disperse_beehler[, 1:2]),
                                directed = FALSE)
plot(my_graph,
     vertex.label = NA, vertex.size = 5)
```

# Discussion/Conclusions

# References
