---
title: "Introduction to ICON"
author: "Raoul R. Wadhwa and Jacob G. Scott"
date: "22 October 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ICON}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Setup

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To prepare for the vignette, we load the necessary R packages and associated datasets.

```{r setup, message=FALSE}
# load R packages (both on CRAN)
library(ICON)
library(network)   # for network analysis
library(ggplot2)   # for network visualization
library(ggnetwork) # for network visualization

# load ICON dataset
data(ICON_data)
```

## Purpose

The ICON R package provides complex networks in edge list format that can be easily incorporated into network analysis pipelines.
The package is based on the [Index of COmplex Networks (ICON) website](https://icon.colorado.edu) that curates complex networks and corresponding summary information (e.g. source, number of nodes \& edges, discipline).
Many publications cite the ICON website, however, since the complex networks in the database exist in distinct formats, the authors of each publication likely had to standardize a large set of networks.
The ICON R package aims to decrease redundant scut work (data formatting) by providing a large number of complex networks from the ICON website in a standard format (network format = edge list; file format = binary R Data with extension .rda).
Currently, the ICON R package provides `r nrow(ICON_data)` complex networks.
As can be reasonably inferred by reading this paragraph, referring to the package as "the ICON R package" can become quite tedious with multiple mentions; as such, the R package will be referred to simply as "ICON" for the remainder of this vignette and the website will be referred to as "the ICON website".

To learn more about ICON, you can visit the associated [website](https://rrrlw.github.io/ICON), [GitHub repo](https://github.com/rrrlw/ICON), and [CRAN page](https://CRAN.R-project.org/package=ICON).

## Which datasets are included in ICON?

The `ICON_data` dataset (loaded in the Setup section) provides a summary of all the datasets that are available through ICON.
Using `head`, we can take a look at the first 6.

```{r peek}
head(ICON_data)
```

Clearly, this is not aesthetic due to the number of columns that `ICON_data` contains.
However, it should provide some comfort that the amount of metadata available within `ICON_data` will likely enable tracking down the original dataset.
To get a nicer view of the available datasets, let's take a look at only a subset of the available metadata.

```{r small_peek}
head(ICON_data[, c("Var_name",
                   "Edges",
                   "Directed",
                   "Name")],
     n = 5)
```

A key difference to note is that the `Var_name` column refers to the dataset name that should be used when accessing it through ICON whereas the `Name` column lists a more descriptive dataset name with little programmatic relevance.
Two salient points should be made here.
First, descriptions of the information contained in each column of `ICON_data` can be accessed in the package documentation via `?ICON_data` or equivalents.
Second, package metadata is generally available in R data packages via package documentation, which avoids the need for a dataset like `ICON_data`.
However, since the networks that ICON provides can be quite large, even in a compressed binary format, they are not hosted within the package on the [Comprehensive R Archive Network (CRAN)](https://CRAN.R-project.org/).
Instead, the desired packages are only downloaded locally when the user instructs ICON to do so via the `get_data` function (explored in the next section).
Although this comes with the disadvantage of not having all datasets immediately available upon installation of ICON, it does save ICON users considerable space if they only wish to use a small subset of the available datasets.

## How to load a specific dataset in ICON?

Once the list of available datasets has been explored, the complex networks can be downloaded using the `get_data` function and the `Var_name` column in `ICON_data`.
For example, the first network in `ICON_data` has `Var_name` set to `aishihik_intensity`, so we download it and peek as follows.

```{r aishihik}
get_data("aishihik_intensity")

head(aishihik_intensity)
```

Every complex network in ICON is provided as an edgelist stored in a data frame.
Each row of the data frame corresponds to a single edge and the first two columns contain the nodes that define the edge; for directed networks, the first column will always be the source (from) and the second column will always be the sink (to).
Note that since only an edgelist is provided, nodes of degree zero will not be included.
Weighted (and some unweighted) networks will contain more than two columns; the additional columns represent either edge weights or other edge attributes.
The `get_data` function can also download multiple datasets at once.
Note that the `get_data`'s `envir` parameter can be modified to select a different environment; by default, the objects will load on the global environment (`.GlobalEnv`).

```{r dl_mult}
get_data(c("coldlake_intensity", "fullerene_c60"))

head(coldlake_intensity)

head(fullerene_c60)
```

`get_data` also lends itself to a simple solution (following code chunk) to download all the networks available through ICON, however, this should be used with caution as there are a large number of them.
The following chunk is not actually run in the vignette.

```{r dl_all, eval=FALSE}
# download all available complex networks
get_data(ICON_data$Var_name)
```

Once downloaded, the complex networks can be stored locally in a binary (e.g. RDA/RData, RDS) or plain-text (CSV, TXT) format; storing it locally removes the reliance on an internet connection for future use.

## How to analyze or visualize networks acquired through ICON?

Although ICON provides the complex networks, it does not provide functionality to analyze or visualize them.
However, the S3 generic `as.network.ICON` is provided to permit use of the [network](https://CRAN.R-project.org/package=network) and [ggnetwork](https://CRAN.R-project.org/package=ggnetwork) R packages for analysis and visualization, respectively.
The following code shows how to generate an object of class `network` using a previously downloaded complex network.

```{r network_obj}
# make sure that the downloaded network has class `ICON`
class(aishihik_intensity)

# coerce to class `network` (`as.network.ICON` is called)
coerced_network <- as.network(aishihik_intensity)

# check if the coerced network has the correct class
class(coerced_network)

# peek at the coerced network (pay attention to number of edges)
coerced_network

# check number of vertices in initial ICON object
length(unique(c(aishihik_intensity[, 1], aishihik_intensity[, 2])))

# check number of edges in initial ICON object
nrow(aishihik_intensity)
```

The initial network (named `aishihik_intensity`) and the coerced network (named `coerced_network`) both contain `r get.network.attribute(coerced_network, "n")` vertices and `r nrow(aishihik_intensity)` edges.
Once we have an object of class `network`, we can visualize it easily using ggnetwork (a ggplot2 extension).

```{r network_viz, fig.width = 4}
# ggnetwork fortifies objects of class `network` without additional code
# the aes parameters should be used as-is
ggplot(coerced_network, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges() +
  geom_nodes() +
  theme_blank()
```

`coerced_network` also has an edge attribute named `Intensity` (see the name of the third column name in `aishihik_intensity`).
This edge attribute can be used to color the edges as follows.

```{r col_edges, fig.width = 5}
ggplot(coerced_network, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(color = Intensity)) + # this line changed
  geom_nodes() +
  theme_blank()
```

Implementing an S3 generic for the `as.network` method and `ICON` class thus makes ICON datasets compatible with the network and ggnetwork packages, considerably simplifying the processes for network analysis and visualization.
ICON's [GitHub README](https://github.com/rrrlw/ICON/blob/master/README.md) provides sample code to analyze and visualize ICON complex networks using [igraph](https://CRAN.R-project.org/package=igraph) as an alternative.